The \dealii Library, Version 9.6

### 2024-09-12 9.6/paper.tex
Work on the paper.

The major changes of this release are:
%
\begin{itemize}
  \item 
\end{itemize}
%

While all of these major changes are discussed in detail in
Section~\ref{sec:major}, there
are a number of other noteworthy changes in the current \dealii release,
which we briefly outline in the remainder of this section:
%
\begin{itemize}
\item changes of interface of AffineConstraints
\item clean up: orientation
\item various performance improvements throughout the library
\item the \texttt{SolverGMRES} class now offers a third orthogonalization
  method, the classical Gram--Schmidt method with delayed orthogonalization
  \cite{Bielich2022}. Furthermore, the solver now specifies the maximal basis
  size of the Arnoldi basis, rather than the number of auxiliary vectors.
\end{itemize}
%
The changelog lists more than X other features and bugfixes.

@@

This release of \dealii contains a number of large and significant changes,
which will be discussed in this section.
It of course also includes a
vast number of smaller changes and added functionality; the details of these
can be found
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_4_2_and_9_5_0.html}
{in the file that lists all changes for this release}; see \cite{changes95}.

\begin{itemize}
\item Hermite element
\item simplices: cubic simplex elements, improved refinement of tetrahedron cells
\item Tpetra advances
\item Nedelec for meshes with hanging nodes
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Updates to matrix-free algorithms}\label{sec:mf}

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interface to the \trilinos \tpetra stack}\label{sec:tpetra}

As a software project, \trilinos{} is designed as a stack of semi-independent
packages building upon each other and working together. 
As a foundation, \teuchos{} offers a set of basic tools,
including parameter handling through XML files or command line interfaces 
and the reference counted pointer \texttt{RCP} which predates 
the smart pointer types introduced in \texttt{C++11}.
The second layer in the stack is a package providing distributed linear algebra 
classes for vectors and (sparse) matrices based on the \petra{}
object model. The third layer is a set of interconnected packages
providing preconditioners, various types of solvers 
as well as more advanced techniques like automatic differentiation, 
time integration and discretization.

For now, there are two choices for the second layer 
with \epetra{} providing \MPI{} only parallelism 
and the newer \tpetra{} (templated Petra) providing additional shared memory 
parallelism and GPU capabilities through \kokkos{}, referred to as \MPIx.
Since the interface of \tpetra{} was overhauled and modernized 
many of the packages in the third layer now have two generations as well.
Therefore, the \MPI-only set of packages is also referred to as the 
\epetra{} stack and the \MPIx{} packages form the \tpetra{} stack.
As a consequence of the different interfaces, 
a decision was made in the \trilinos{} project to deprecate the \epetra{} stack 
and only provide basic maintenance until a full 
removal from their repository in the course of 2025.

\begin{itemize}
  \item Short overview capabilities of \epetra (MPI parallel, fixed types) [Thiele]
  \item Short overview capabilities of \tpetra (MPI+X via Kokkos, auto diff types, Templated petra) [Thiele]
  \item Current state of the interface (Tpertra based linear algebra, direct solver from Amesos2, preconditioners from Ifpack2) [Thiele \& Kinnewig]
  \item User relevant changes from \epetra: incompatibilities, removed options, changes in the constructor. [Thiele \& Kinnewig]
\end{itemize}

@@

Furthermore, we added an example to the \texttt{libCEED}
library\footnote{\url{https://github.com/CEED/libCEED}}. \texttt{libCEED} is a library
that provides matrix-free evaluation routines that work on different hardware. The
examples show how to interface the \dealii data structures to the \texttt{libCEED} ones.
The BP1-BP6 benchmarks (scalar/vector Laplace/mass matrix are solved with
regular integration and over-integration) are solved.

@@

The 9.6 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_5_2_and_9_6_0.html}
     {around X incompatible changes};
see \cite{changes96}. Many of these
incompatibilities change internal
interfaces that are not usually used in external
applications. That said, the following are worth mentioning since they
may have been more widely used:
\begin{itemize}
  \item CUDAWrappers, that enable the usage of cuSPARSE algorithms, have been deprecated and will be removed in the next release. Kokkos is now used for device-specific optimizations.
\end{itemize}

>>>>

The major changes of this release are:
%
\begin{itemize}
  \item Substantial improvements to the multigrid two-level transfer
    operators, including matrix-free based versions and versions for
    p-refinement (Section~\ref{sec:mf}).
  \item A class FERemoteEvaluation that provides support for evaluating
    finite element shape functions and solutions on parts of the
    domain stored by other MPI processes (Section~\ref{sec:remote-eval}).
  \item Much work has gone into writing wrappers for Trilinos' Tpetra
    stack of linear algebra classes. Tpetra is Trilinos' Kokkos-based
    replacement for the now-deprecated Epetra stack. See
    Section~\ref{sec:tpetra} for more on this.
  \item Tool classes \texttt{TaskResult} and \texttt{Lazy} that provide ways to compute
    values on a separate thread or only when first accessed (Section~\ref{sec:tools}).
  \item There are six new tutorial programs, on checkpointing simulations
    (step-83), integrating time-dependent solvers with external time
    stepping libraries (step-86, using PETSc's TS library), advanced
    point evaluation techniques (step-87), non-matching grids
    (step-89), and trace-based methods for PDEs on embedded surfaces
    (step-90). See Section~\ref{subsec:steps} for more details.
\end{itemize}
%

\todo[inline]{If we keep the section on relicensing
  (Section~\ref{sec:license}), mention it in the list above.}

While all of these major changes are discussed in detail in
Section~\ref{sec:major}, there
are a number of other noteworthy changes in the current \dealii release,
which we briefly outline in the remainder of this section:
%
\begin{itemize}
  \item deal.II now requires and uses of C++17.
  \item We have continued to make progress in supporting simplex and
    mixed meshes -- mesh types that \dealii{} has traditionally not
    supported at all. Specifically, the current release uses
    better strategies for refinement of tetrahedra that result in
    better-shaped child cells. It also contains
    support for cubic finite elements on simplices.
  \item The \texttt{FE\_NedelecSZ} class that contains our
    implementation of the N\'ed\'elec element using the orientation
    scheme of \cite{Zag06} now supports the computation of hanging
    node constraints for locally refined meshes.
    \todo[inline]{Sebastian: Is there anything more to say here? If
      you want, we can also list this point in the list above, and add
      a whole section 2.x for it.}
  \item changes of interface of AffineConstraints
    \todo[inline]{Wolfgang to write}
\item clean up: orientation
  \todo[inline]{DavidW: Please describe some details here.}
\item Various performance improvements throughout the library.
  \todo[inline]{Everyone: This sentence would be more useful if we can
    provide a few examples.}
\item The \texttt{SolverGMRES} class now offers a third orthogonalization
  method, the classical Gram--Schmidt method with delayed orthogonalization
  \cite{Bielich2022}. Furthermore, the solver now specifies the maximal basis
  size of the Arnoldi basis, rather than the number of auxiliary
  vectors. Some changes have also been made to the GMRES and F-GMRES
  implementations to make them more similar.
  \todo[inline]{Martin: Does this capture the relevant changes? Please
    fix as appropriate, or if you'd like put them into a section of
    its own.}
\end{itemize}
%
The changelog lists more than 180 other features and bugfixes.

@@

This release of \dealii contains a number of large and significant changes,
which we will discuss in this section.
It of course also includes a
vast number of smaller changes and added functionality; the details of these
can be found
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_5_2_and_9_6_0.html}
{in the file that lists all changes for this release}; see \cite{changes96}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Updates to multigrid and matrix-free algorithms}\label{sec:mf}

\todo[inline]{Martin/Peter: The list in the introduction also mentions the two-level
  operators. Is this the right place to also discuss these?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The \texttt{FERemoteEvaluation} class}\label{sec:remote-eval}

\todo[inline]{Peter/Martin/...: Would you be willing to describe this
  class and its uses? Or, if you don't think it's important enough,
  remove this section and move the part in the top list of major
  changes to the bottom list?}

@@

\todo[inline]{All: If you contributed to this functionality, please
  help flesh out this section.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interface to the \trilinos{} \tpetra stack}\label{sec:tpetra}

\dealii{}'s parallel linear algebra facilities are largely built on
wrappers around functionality provided by the \petsc{} and \trilinos{} libraries.
As a software project, \trilinos{} is designed as a stack of semi-independent
packages building upon each other and working together. 
As a foundation, \teuchos{} offers a set of basic tools,
including parameter handling through XML files or command line interfaces 
and the reference counted pointer \texttt{RCP} which predates 
the smart pointer types introduced in \texttt{C++11}.
The second layer in the stack is a package providing distributed linear algebra 
classes for vectors and (sparse) matrices based on the \petra{}
object model. The third layer is a set of interconnected packages
providing preconditioners, various types of solvers 
as well as more advanced techniques like automatic differentiation, 
time integration and discretization.

Historically, \trilinos{} has implemented \petra{} objects -- such as
parallel vector, sparsity pattern, and sparse matrix classes -- in its
\epetra{} package that uses \MPI{} as its only source of
parallelism. The \dealii{} interfaces to \trilinos{} are therefore all
built using \epetra{}.
However, several years ago, \trilinos also introduced the
newer \tpetra{} (templated Petra) package that provides additional shared memory 
parallelism and GPU capabilities by building on \kokkos{}. \tpetra{}
is slated to eventually replace \epetra{}, and the latter is indeed
now deprecated. As a consequence, we will eventually have to switch
all of our \trilinos{} interfaces to \tpetra{} and \tpetra{}-based sub-packages.

In the current release, we have put substantial work into this switch.
Since the interface of \tpetra{} was overhauled and modernized 
many of the packages in the third layer now have two generations as well.
Therefore, the \MPI-only set of packages is also referred to as the 
\epetra{} stack and the \MPIx{} packages form the \tpetra{} stack.
As a consequence of the different interfaces, 
a decision was made in the \trilinos{} project to deprecate the \epetra{} stack 
and only provide basic maintenance until a full 
removal from their repository in the course of 2025.

\begin{itemize}
  \item Short overview capabilities of \epetra (MPI parallel, fixed types) [Thiele]
  \item Short overview capabilities of \tpetra (MPI+X via Kokkos, auto diff types, Templated petra) [Thiele]
  \item Current state of the interface (Tpetra based linear algebra, direct solver from Amesos2, preconditioners from Ifpack2) [Thiele \& Kinnewig]
  \item User relevant changes from \epetra: incompatibilities, removed options, changes in the constructor. [Thiele \& Kinnewig]
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{More support for advanced programming idioms}\label{sec:tools}

\todo[inline]{Wolfgang to write. Mention Lazy and TaskResult.}

@@

Furthermore, we added an example to the \texttt{libCEED}
library.\footnote{\url{https://github.com/CEED/libCEED}} \texttt{libCEED} is a library
that provides matrix-free evaluation routines that work on different hardware. The
examples show how to interface the \dealii data structures to the \texttt{libCEED} ones.
The example solves the BP1-BP6 benchmarks (scalar/vector Laplace/mass matrix with
regular integration and over-integration).

\todo[inline]{Matthias: Where are we with this, and what of it is
  worth describing here?}

@@

The 9.6 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_5_2_and_9_6_0.html}
     {around 40 incompatible changes};
see \cite{changes96}. Many of these
incompatibilities change internal
interfaces that are not usually used in external
applications. That said, the following are worth mentioning since they
may have been more widely used:
\begin{itemize}
  \item \dealii{} now requires compilers to support C++17, and has
    started to extensively use C++17 features.
  \item The \texttt{CUDAWrappers} namespace and its contents --
    notably things that enable the usage of cuSPARSE algorithms --
    have been deprecated and will be removed in the next
    release. Kokkos is now used for device-specific optimizations.
\end{itemize}
